---
title: "Math of mortality"
description: |
  Post on mathematical identities used in mortality estimation. STAN codes to simulate survival data and to estimate Piecewise Constant Hazards model.
author:
  - name: Benjamin Schlüter
date: 02-10-2022
output:
  distill::distill_article:
    self_contained: false
    highlight: monochrome
draft: TRUE
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Survival analyses can be analyzed (at minimal) with three complementary functions: hazard, survival and death probability density function.

Let denote X: the age at death of an individual and assume that this random variable is $X\geq 0$ and continuous. 

Suriving beyond age x is expressed as

$$
\begin{align}
S(x) & = Pr(X>x) \\
S(x) & = 1 - F(x) \\
S'(x) & = -f(x) 
\end{align}
$$

$$S(x) = Pr(X > x) = -\int^{\infty}_xf(t)dt$$

Let's now express the hazard function, also called the force of mortality:

$$
\begin{align}
h(x) &= lim_{\Delta x \rightarrow 0} \frac{Pr(x \leq X < x + \Delta x | X > x)}{\Delta x} \\
 &= lim_{\Delta x \rightarrow 0} \frac{Pr(X < x + \Delta x) - Pr(X \leq x)}{\Delta x} \cdot \frac{1}{Pr(X>x)} \\
 &= \frac{f(t)}{S(t)}
\end{align}
$$

Thus, the force of mortality can be expressed as

$$
\begin{align}
\boxed{h(x) = -\frac{d}{dx}ln(S(x))}
\end{align}
$$

and re-arranging that equation we express the survival probability at age x as

$$\boxed{S(x) = e^{-\int^x_0h(t)dt} = e^{-H(x)}}$$

where $H(t)$ is called the cumulative hazard.

We can further express $S(x+n)$ as

$$S(x+n) = S(x)e^{-\int^{x+n}_{x}h(t)dt}$$



Note that when $h(x)$ is constant, 

$$S(t) = e^{-ht}$$

From earlier derivation of $h(x)$ we can obtain a math expression for the death probability density function

$$\boxed{f(x) = h(x)S(x)}$$

Let's check these identities visually or model a survival model with STAN ?


```{r obtain_a_b, eval = FALSE, include = FALSE}

# ONLY PURPOSE IS TO GET A AND B VALUES THAT MAKE SENSE
# BEL 2015, 55-95 yo estimated with poisson regression

library(ggplot2)
library(tidyverse)
# Select a and b according to BEL data in 2015
dth = read.delim("./modeling/data/dthBE_1x1.txt", 
                 skip = 1,
                 sep = "", 
                 stringsAsFactors=FALSE,
                 header = TRUE)

exp = read.delim("./modeling/data/expBE_1x1.txt", 
                 skip = 1,
                 sep = "", 
                 stringsAsFactors=FALSE,
                 header = TRUE)

df.dth = dth %>% 
  filter(Year == 2015) %>% 
  select(Year, Age, Female) %>% 
  mutate(Female = as.numeric(Female))

df.exp = exp %>% 
  filter(Year == 2015) %>% 
  select(Year, Age, Female)

df = df.dth %>% 
  left_join(df.exp, 
            by = c("Year", "Age")) %>% 
  rename("D" = Female.x,
         "N" = Female.y) %>% 
  mutate(mx = D/N,
         Age = ifelse(Age == "110+", 110, Age),
         Age = as.numeric(Age))
# Gompertz perfom badly for young and old ages
# Age bounds
age.bds <- c(55, 95)
df %>% 
  filter(Age >= age.bds[1], 
         Age < age.bds[2]) %>% 
  ggplot(aes(x = Age, y = log(mx))) +
  geom_point()

# create data from age 55
df <- df %>% 
  filter(Age >= age.bds[1], 
         Age < age.bds[2]) %>% 
  mutate(age = Age - age.bds[1])

# important that age start at 0
fit.gomp <- lm(log(mx) ~ 1 + age,
               data = df )
fit.gomp$coef

df %>% 
  mutate(fit = fit.gomp$coef[1] + fit.gomp$coef[2]*age) %>% 
  ggplot(aes(x = age, y = log(mx))) +
  geom_point() +
  geom_line(aes(x = age, y = fit))

a <- exp(fit.gomp$coef[1])
b <- fit.gomp$coef[2]

df %>% 
  mutate(fit = a * exp(b*age)) %>% 
  ggplot(aes(x = age, y = mx)) +
  geom_point() +
  geom_line(aes(x = age, y = fit))

```


```{r stan_sim, eval = FALSE, include = FALSE}
library(rstan)

writeLines(readLines("./modeling/stan/simu_gomp.stan"))

# sample size
N <- 2e4

stan_data <- list("N" = N,
                  "a" = a,
                  "b" = b)

# simu <- stan(file="./modeling/stan/simu_gomp.stan", data=stan_data,
#              iter=1, chains=1, seed=411991,
#              algorithm="Fixed_param")
# 
# t <- array(extract(simu)$t[1,])
# range(t)



# Much quicker without STAN
u <- runif(N, 0, 1)
t <- (1/b)*log(1-((b/a)*log(1-u))) # t >= 0 as age start at 85 thanks to par a

```


```{r, echo = FALSE}

# Comparison between truth (Gompertz) and simulation

# let's first plot the true density of yi
grid <- seq(0, ceiling(max(t)), 0.1)
df <- tibble(x = grid,
             h = a*exp(b*grid),
             S = exp((a/b)*(1-exp(b*grid))),
             pi = h*S)

# hazard
ggplot(df, aes(x = x, y = h)) +
  geom_line() +
  labs(y = "h(x)")

# density
ggplot(df, aes(x = x, y = pi)) +
  geom_line()

# compare histogram with true pdf of t
ggplot() +
  geom_histogram(data = tibble(x = t), 
                 aes(x = x, y = ..density..), 
                 breaks = seq(0, ceiling(max(t)), 0.1)) +
  geom_line(data = df, 
            aes(x = x, y = pi),
            col = "red")
```


```{r, echo = FALSE}

# WITHOUT ANY TRUNCATION HOW DOES POISSON PERFORM?
# eps <- 1
# dth <- c()
# exp <- c()
# 
# brks <- seq(0, ceiling(max(t)), eps)
# for (i in 1:(length(brks)-1)){
#     # dth
#     d <- sum( (t >= brks[i] & t < brks[i+1]) )
#     # time lived by those dying
#     e_d <- t[t >= brks[i] & t < brks[i+1]] - brks[i]
#     e <- sum(t >= brks[i])*eps + sum(e_d)
#   dth <- c(dth, d)
#   exp <- c(exp, e)
# }

# Nice way to get count data (DHS)
# Easy as interval is only 1 year
df.counts <- tibble(t = t) %>% 
  # obtain lower bound
  mutate(interval = floor(t),
         # exposure from those dying
         exp_d = t - interval) %>% 
  group_by(interval) %>% 
  summarise(dth = n(),
            exp_d = sum(exp_d)) %>% 
  ungroup() %>% 
  mutate(mid = interval + .5,
         dth.cum = cumsum(dth),
         # exposure from those surviving
         exp_s = N - dth.cum,
         exp = exp_d + exp_s,
         mx = dth/exp)

ggplot(df.counts, aes(x = mid, y = mx)) +
  geom_point()

# # df raw estimates
# df.poi <- tibble(cut_off = head(brks, -1) + (eps/2), # mid of interval
#                  dth = dth,
#                  exp = exp) %>%
#   mutate(mx =dth/exp)
# 
# # Fit poisson  
# brks <- head(brks, -1) + (eps/2) # mid of interval
# dth <- dth[exp > 0]
# brks <- brks[exp > 0]
# exp <- exp[exp > 0]
fit.poi <- glm(dth ~ 1 + mid, offset = log(exp),
           family = "poisson",
           data = df.counts)

# Compare fit with obtained raw estimated from simulated data
df.counts <- df.counts %>%
  mutate(fit.glm = exp(fit.poi$coefficients[1])*exp(fit.poi$coefficients[2]*mid) )

# Compare fit with truth
ggplot() +
  geom_line(data = df, aes(x = x, y = h, col = "truth")) +
  geom_line(data = df.counts, aes(x = mid, y = fit.glm, col = "Poisson")) +
  geom_point(data = df.counts, aes(x = mid, y = mx, col = "nmx simu"),
             alpha = 0.1) +
  scale_color_manual(values = c("truth"=  "skyblue3",
                                  "Poisson" = "slateblue",
                                  "nmx simu" = "darkseagreen")) +
  labs(y = "h(x)")

# Do we find the parameter back ? YES
abs(exp(fit.poi$coefficients[1]) - a)
abs(fit.poi$coefficients[2] - b)
```


```{r, eval = FALSE, include = FALSE}
 # Poisson estimation with STAN
stan_data <- list(
  "A" = dim(df.counts)[1],
  "age" = df.counts$mid,
  "n" = df.counts$exp,
  "d" = df.counts$dth
  )

options(mc.cores = parallel::detectCores()-1)

fit = stan("./modeling/stan/poi_for_gomp.stan",
           iter = 8000,
           chains = 4,
           data = stan_data)

library(tidybayes)

df.fit.ab <- fit %>% 
  spread_draws(a, b) %>% 
  median_qi() 

df.fit <- fit %>% 
  spread_draws(eta[i]) %>% 
  median_qi() %>% 
  mutate(mid = i-0.5,
         mx = exp(eta)) 

# Comparison GLM, STAN and truth
ggplot() +
  geom_line(data = df.fit, aes(x = mid, y = mx)) +
  geom_line(data = df.counts, aes(x = mid, y = fit.glm), col = "red", linetype = "dashed") +
    geom_line(data = df, aes(x = x, y = h), col = "skyblue3", size = 1.5,
              alpha = 0.3)
  
# Obtain S and pi from estimated mx
a.hat <- exp(extract(fit)$a)
b.hat <- extract(fit)$b

h.hat = sapply( grid, function(x) a.hat*exp(b.hat*x) )
S.hat = sapply( grid, function(x) exp((a.hat/b.hat)*(1-exp(b.hat*x))) )
pi.hat = h.hat*S.hat

df.hat <- tibble(x = grid,
             h = apply(h.hat, 2, median),
             h.low = apply(h.hat, 2, quantile, probs = 0.025),
             h.up = apply(h.hat, 2, quantile, probs = 0.975),
             S = apply(S.hat, 2, median),
             S.low = apply(S.hat, 2, quantile, probs = 0.025),
             S.up = apply(S.hat, 2, quantile, probs = 0.975),
             pi = apply(pi.hat, 2, median),
             pi.low = apply(pi.hat, 2, quantile, probs = 0.025),
             pi.up = apply(pi.hat, 2, quantile, probs = 0.975))

# Compare h with true
ggplot() +
  geom_line(data = df.hat, aes(x = x, y = h),
            col = "skyblue3") +
  geom_ribbon(data = df.hat, aes(x = x, ymin = h.low, ymax = h.up),
            col = "skyblue3",
            fill = "skyblue3",
            alpha = 0.3) +
  geom_line(data = df, aes(x = x, y = h), 
            size = 1.1)


# Compare S with true
ggplot() +
  geom_line(data = df.hat, aes(x = x, y = S),
            col = "skyblue3") +
  geom_ribbon(data = df.hat, aes(x = x, ymin = S.low, ymax = S.up), 
              col = "skyblue3",
              fill = "skyblue3",
              alpha = 0.3) +
  geom_line(data = df, aes(x = x, y = S), 
            size = 1.2)

# Compare pi with true
ggplot() +
  geom_line(data = df.hat, aes(x = x, y = pi),
            col = "skyblue3") +
  geom_ribbon(data = df.hat, aes(x = x, ymin = pi.low, ymax = pi.up),
              col = "skyblue3",
              fill = "skyblue3",
              alpha = 0.3) +
  geom_line(data = df, aes(x = x, y = pi), 
            size = 1.2)


```


```{r, echo = FALSE}

# Check the effect of right censoring


# censoring time with ICDF from exponential
set.seed(2841992)
u <- runif(N, 0, 1)
nu <- 1.5e-2 # strength of censoring, higher more censoring. 
ci <- -log(1-u)/nu

# observed exit time
gamma <- t > ci
yi <- pmin(t, ci)
# sum(gamma == TRUE)/N: % whose death are observed

# need to account for censoring (not everybody is seen dying)
# still work as dying has same effect as being censored
df.counts <- tibble(t = t,
                    cens = gamma,
                    y = yi) %>% 
  # obtain lower bound
  mutate(interval = floor(y),
         # exposure from those exiting (dth/censored)
         exp_out = (y - interval)) %>% 
  group_by(interval, cens) %>% 
  summarise(dth = n(),
            exp_out = sum(exp_out)) %>% 
  ungroup() %>% 
  # count people exiting at each interval
  group_by(interval) %>% 
  mutate(out = sum(dth),
         exp_out = sum(exp_out)) %>% 
  ungroup() %>% 
  # account the censored obs
  # only through exposure
  filter(cens == FALSE) %>% 
  mutate(mid = interval + .5,
         out.cum = cumsum(out),
         # exposure from those surviving/not censored
         exp_s = N - out.cum,
         exp = exp_out + exp_s,
         mx = dth/exp)


# Poisson estimation with STAN
stan_data <- list(
  "A" = dim(df.counts)[1],
  "age" = df.counts$mid,
  "n" = df.counts$exp,
  "d" = df.counts$dth
  )

options(mc.cores = parallel::detectCores()-1)

fit = stan("./modeling/stan/poi_for_gomp.stan",
           iter = 8000,
           chains = 4,
           data = stan_data)


# Obtain S and pi from estimated mx
a.hat <- exp(extract(fit)$a)
b.hat <- extract(fit)$b

h.hat = sapply( grid, function(x) a.hat*exp(b.hat*x) )
S.hat = sapply( grid, function(x) exp((a.hat/b.hat)*(1-exp(b.hat*x))) )
pi.hat = h.hat*S.hat

df.hat <- tibble(x = grid,
             h = apply(h.hat, 2, median),
             h.low = apply(h.hat, 2, quantile, probs = 0.025),
             h.up = apply(h.hat, 2, quantile, probs = 0.975),
             S = apply(S.hat, 2, median),
             S.low = apply(S.hat, 2, quantile, probs = 0.025),
             S.up = apply(S.hat, 2, quantile, probs = 0.975),
             pi = apply(pi.hat, 2, median),
             pi.low = apply(pi.hat, 2, quantile, probs = 0.025),
             pi.up = apply(pi.hat, 2, quantile, probs = 0.975))

# Compare h with true
ggplot() +
  geom_line(data = df.hat, aes(x = x, y = h, col = "true")) +
  geom_ribbon(data = df.hat, aes(x = x, ymin = h.low, ymax = h.up,
                                 col = "estimated",
                                 fill = "estimated"),
              alpha = 0.3) +
  geom_line(data = df, aes(x = x, y = h), 
            size = 1.05) +
  scale_color_manual(values = c("true" = "black",
                            "estimated" = "skyblue3")) +
  scale_fill_manual(values = c("true" = "black",
                               "estimated" = "skyblue3")) +
  scale_y_log10()


# Compare S with true
ggplot() +
  geom_line(data = df.hat, aes(x = x, y = S, col = "true")) +
  geom_ribbon(data = df.hat, aes(x = x, ymin = S.low, ymax = S.up,
                                 col = "estimated",
                                 fill = "estimated"),
              alpha = 0.3) +
  geom_line(data = df, aes(x = x, y = S), 
            size = 1.05) +
  scale_color_manual(values = c("true" = "black",
                            "estimated" = "skyblue3")) +
  scale_fill_manual(values = c("true" = "black",
                               "estimated" = "skyblue3"))

# Compare pi with true
ggplot() +
  geom_line(data = df.hat, aes(x = x, y = pi, col = "true")) +
  geom_ribbon(data = df.hat, aes(x = x, ymin = pi.low, ymax = pi.up,
                                 col = "estimated",
                                 fill = "estimated"),
              alpha = 0.3) +
  geom_line(data = df, aes(x = x, y = pi), 
            size = 1.05) +
  scale_color_manual(values = c("true" = "black",
                            "estimated" = "skyblue3")) +
  scale_fill_manual(values = c("true" = "black",
                               "estimated" = "skyblue3"))


```

















```{r simulation, eval = FALSE, include = FALSE}

# CAMARDA SIMULATION SET-UP


# By setting nu really low and d0 to b0 there is not truncation to test estimation

# entry time, when an individual reaches 85 yo: b_0=0 ; b_max=10
bi <- runif(N, 0, 10)

# censoring time with ICDF from exponential
u <- runif(N, 0, 1)
nu <- 3e-2 # strength of censoring, higher more censoring. 
ci <- -log(1-u)/nu

# timing of death wit ICDF of gompertz
# a and b define with real data
# --> model in STAN
# u <- runif(N, 0, 1)
# ti <- (1/b)*log(1-((b/a)*log(1-u))) # t >= 0 as age start at 85 thanks to par a

# observed exit time
gamma <- t < ci
yi <- pmin(t, ci)
# sum(gamma == TRUE)/N: % whose death are observed

# date of exit when an individual reaches 85 yo (bi) + age at death/censoring (yi)
di <- bi + yi 

# beginning of follow-up 
d0 <- 0

# left-truncation: individuals exits before the follow-up
# new sample
yi <- yi[di>d0]

# enter before d0 and survive start of the follow-up: wi is entry age (85+wi)
wi <- d0 - bi

# individual entering after d0, enter age 105
wi[wi<0] <- 0

```


```{r estimation, eval = FALSE, include = FALSE}
# does yi look like gompertz ?

# let's first plot the true density of yi
grid <- seq(0, ceiling(max(yi)), 0.05)
df <- tibble(x = grid,
             h = a*exp(b*grid),
             S = exp((a/b)*(1-exp(b*grid))),
             pi = h*S)

# hazard
ggplot(df, aes(x = x, y = h)) +
  geom_line() +
  labs(y = "h(x)")

# density
ggplot(df, aes(x = x, y = pi)) +
  geom_line()

# compare with histogram
ggplot() +
  geom_histogram(data = tibble(x = yi), 
                 aes(x = x, y = ..density..), 
                 breaks = seq(0, ceiling(max(yi)), 0.05)) +
  geom_line(data = df, 
            aes(x = x, y = pi),
            col = "red")

# see if capable to estimate with Poisson model
eps <- 1
dth <- c()
exp <- c()

brks <- seq(0, ceiling(max(yi)), eps)
for (i in 1:(length(brks)-1)){
    # dth
    d <- sum( (yi >= brks[i] & yi < brks[i+1]) )
    # time lived by those dying
    e_d <- yi[yi >= brks[i] & yi < brks[i+1]] - brks[i]
    e <- sum(yi >= brks[i])*eps + sum(e_d)
  dth <- c(dth, d)
  exp <- c(exp, e)
  
}

# df raw estimates
df.poi <- tibble(cut_off = head(brks, -1) + (eps/2), # mid of interval
                 dth = dth,
                 exp = exp) %>%
  mutate(mx =dth/exp)

ggplot() +
  geom_point(data = df.poi, aes(x = cut_off, y = mx))

# Fit poisson  
brks <- head(brks, -1) + (eps/2) # mid of interval
dth <- dth[exp > 0]
brks <- brks[exp > 0]
exp <- exp[exp > 0]
fit.poi <- glm(dth ~ 1 + brks, offset = log(exp),
           family = "poisson")

# Compare fit with obtained raw estimated from simulated data
df.poi <- df.poi %>%
  mutate(fit = exp(fit.poi$coefficients[1])*exp(fit.poi$coefficients[2]*cut_off) )

ggplot(df.poi) +
  geom_point(aes(x = cut_off, y = mx)) +
  geom_line(aes(x = cut_off, y = fit))

# Compare fit with truth
ggplot() +
  geom_line(data = df, aes(x = x, y = h, col = "truth")) +
  geom_line(data = df.poi, aes(x = cut_off, y = fit, col = "Poisson")) +
  geom_point(data = df.poi, aes(x = cut_off, y = mx, col = "nmx simu"),
             alpha = 0.3) +
  scale_color_manual(values = c("truth"=  "skyblue3",
                                  "Poisson" = "slateblue",
                                  "nmx simu" = "darkseagreen")) +
  scale_y_log10() +
  labs(y = "h(x)")

# Do we find the parameter back ? YES
abs(exp(fit.poi$coefficients[1]) - a)
abs(fit.poi$coefficients[2] - b)
```




