---
title: "Math of mortality"
description: |
  Post showing the derivations of some mathematical identities used in mortality estimation with a simulation and estimation in STAN.
author:
  - name: Benjamin Schlüter
date: 02-10-2022
output:
  distill::distill_article:
    self_contained: false
    highlight: monochrome
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

The field of mortality can be analyzed (at minimal) with three complementary functions: hazard, survival and death probability density function.

Let denote X: the age at death of an individual and assume that this random variable is $X\geq 0$ and continuous. 

Suriving beyond age x is expressed as

$$
\begin{align}
S(x) & = Pr(X>x) \\
S(x) & = 1 - F(x) \\
S'(x) & = -f(x) 
\end{align}
$$

$$S(x) = Pr(X > x) = -\int^{\infty}_xf(t)dt$$

Let's now express the hazard function, also called the force of mortality:

$$
\begin{align}
h(x) &= lim_{\Delta x \rightarrow 0} \frac{Pr(x \leq X < x + \Delta x | X > x)}{\Delta x} \\
 &= lim_{\Delta x \rightarrow 0} \frac{Pr(X < x + \Delta x) - Pr(X \leq x)}{\Delta x} \cdot \frac{1}{Pr(X>x)} \\
 &= \frac{f(t)}{S(t)}
\end{align}
$$

Thus, the force of mortality can be expressed as

$$
\begin{align}
\boxed{h(x) = -\frac{d}{dx}ln(S(x))}
\end{align}
$$

and re-arranging that equation we express the survival probability at age x as

$$\boxed{S(x) = e^{-\int^x_0h(t)dt} = e^{-H(x)}}$$

where $H(t)$ is called the cumulative hazard.

We can further express $S(x+n)$ as

$$S(x+n) = S(x)e^{-\int^{x+n}_{x}h(t)dt}$$



Note that when $h(x)$ is constant, 

$$S(t) = e^{-ht}$$

From earlier derivation of $h(x)$ we can obtain a math expression for the death probability density function

$$\boxed{f(x) = h(x)S(x)}$$

Let's check these identities visually or model a survival model with STAN ?


```{r obtain_a_b, eval = FALSE, include = FALSE}
library(ggplot2)
library(tidyverse)
# Select a and b according to BEL data in 2015
dth = read.delim("./modeling/data/dthBE_1x1.txt", 
                 skip = 1,
                 sep = "", 
                 stringsAsFactors=FALSE,
                 header = TRUE)

exp = read.delim("./modeling/data/expBE_1x1.txt", 
                 skip = 1,
                 sep = "", 
                 stringsAsFactors=FALSE,
                 header = TRUE)

df.dth = dth %>% 
  filter(Year == 2015) %>% 
  select(Year, Age, Female) %>% 
  mutate(Female = as.numeric(Female))

df.exp = exp %>% 
  filter(Year == 2015) %>% 
  select(Year, Age, Female)

df = df.dth %>% 
  left_join(df.exp, 
            by = c("Year", "Age")) %>% 
  rename("D" = Female.x,
         "N" = Female.y) %>% 
  mutate(mx = D/N,
         Age = ifelse(Age == "110+", 110, Age),
         Age = as.numeric(Age))

df %>% 
  filter(Age >= 85, 
         Age < 100) %>% 
  ggplot(aes(x = Age, y = mx)) +
  geom_line()
# mx close to 0.02 at 85 yo

# create data from age 85
df <- df %>% 
  filter(Age >= 85, 
         Age < 100) %>% 
  mutate(age = Age - 85)

# important that age start at 0
fit.gomp <- lm(log(mx) ~ 1 + age,
               data = df )
fit.gomp$coef

df %>% 
  mutate(fit = fit.gomp$coef[1] + fit.gomp$coef[2]*age) %>% 
  ggplot(aes(x = age, y = log(mx))) +
  geom_line() +
  geom_line(aes(x = age, y = fit))

a <- exp(fit.gomp$coef[1])
b <- fit.gomp$coef[2]

df %>% 
  mutate(fit = a * exp(b*age)) %>% 
  ggplot(aes(x = age, y = mx)) +
  geom_line() +
  geom_line(aes(x = age, y = fit))

```


```{r stan_sim, eval = FALSE, include = FALSE}
library(rstan)

writeLines(readLines("./modeling/stan/simu_gomp.stan"))

# sample size
N <- 2e4

stan_data <- list("N" = N,
                  "a" = a,
                  "b" = b)

simu <- stan(file="./modeling/stan/simu_gomp.stan", data=stan_data,
             iter=1, chains=1, seed=411991,
             algorithm="Fixed_param")

t <- array(extract(simu)$t[1,])

```



```{r simulation, eval = FALSE, include = FALSE}

# By setting nu really low and d0 to b0 there is not truncation to test estimation

# entry time, when an individual reaches 85 yo: b_0=0 ; b_max=10
bi <- runif(N, 0, 10)

# censoring time with ICDF from exponential
u <- runif(N, 0, 1)
nu <- 1e-6 # strength of censoring, higher more censoring. 
ci <- -log(1-u)/nu

# timing of death wit ICDF of gompertz
# a and b define with real data
# --> model in STAN
# u <- runif(N, 0, 1)
# ti <- (1/b)*log(1-((b/a)*log(1-u))) # t >= 0 as age start at 85 thanks to par a

# observed exit time
gamma <- t < ci
yi <- pmin(t, ci)
# sum(gamma == TRUE)/N: % whose death are observed

# date of exit when an individual reaches 85 yo (bi) + age at death/censoring (yi)
di <- bi + yi 

# beginning of follow-up 
d0 <- 0

# left-truncation: individuals exits before the follow-up
# new sample
yi <- yi[di>d0]

# enter before d0 and survive start of the follow-up: wi is entry age (105+wi)
wi <- d0 - bi

# individual entering after d0, enter age 105
wi[wi<0] <- 0

```


```{r estimation, eval = FALSE, include = FALSE}
# does yi look like gompertz ?

# let's first plot the true density of yi
grid <- seq(0, ceiling(max(yi)), 0.05)
df <- tibble(x = grid,
             h = a*exp(b*grid),
             S = exp((a/b)*(1-exp(b*grid))),
             pi = h*S)

# hazard
ggplot(df, aes(x = x, y = log(h))) +
  geom_line() +
  labs(y = "ln(h(x))")

# density
ggplot(df, aes(x = x, y = pi)) +
  geom_line()

# compare with histogram
ggplot() +
  geom_histogram(data = tibble(x = yi), 
                 aes(x = x, y = ..density..), 
                 breaks = seq(0, ceiling(max(yi)), 0.05)) +
  geom_line(data = df, 
            aes(x = x, y = pi),
            col = "red")

# see if capable to estimate with Poisson model
eps <- 0.05
dth <- c()
exp <- c()

brks <- seq(0, ceiling(max(yi)), eps)
for (i in 1:(length(brks)-1)){
    # dth
    d <- sum( (yi >= brks[i] & yi < brks[i+1]) )
    # time lived by those dying
    e_d <- yi[yi >= brks[i] & yi < brks[i+1]] - brks[i]
    e <- sum(yi >= brks[i])*eps + sum(e_d)
  dth <- c(dth, d)
  exp <- c(exp, e)
  
}

# df raw estimates
df.poi <- tibble(cut_off = head(brks, -1),
                 dth = dth,
                 exp = exp) %>%
  mutate(mx =dth/exp)

ggplot() +
  geom_point(data = df.poi, aes(x = cut_off, y = mx))
  
brks <- head(brks, -1)
dth <- dth[exp > 0]
brks <- brks[exp > 0]
exp <- exp[exp > 0]
fit <- glm(dth ~ 1 + brks, offset = log(exp),
           family = "poisson")

ggplot(data = df.poi %>% 
         mutate(fit = exp(fit$coefficients[1])*exp(fit$coefficients[2]*cut_off) )) +
  geom_point(aes(x = cut_off, y = mx)) +
  geom_line(aes(x = cut_off, y = fit))

# Do we find the parameter back ? YES
abs(exp(fit$coefficients[1]) - a)
abs(fit$coefficients[2] - b)
```



```{r, eval = FALSE, include = FALSE}
 # Poisson estimation with STAN
stan_data <- list(
  "A" = length(brks),
  "age" = brks,
  "n" = round(exp,0),
  "d" = dth
  )

options(mc.cores = parallel::detectCores()-1)

fit = stan("./modeling/stan/poi_for_gomp.stan",
           iter = 4000,
           data = stan_data)

library(tidybayes)

df.fit.ab <- fit %>% 
  spread_draws(a, b) %>% 
  median_qi() 

df.fit <- fit %>% 
  spread_draws(eta[i]) %>% 
  median_qi() %>% 
  mutate(age = i-1,
         mx = exp(eta),
         # check pdf of t when assuming d~Poisson
         pi = mx*exp(-mx*age)) 

ggplot() +
  geom_line(data = df.fit, aes(x = age, y = eta))
# pdf of t when assuming d~poisson
# and compare with truth
ggplot() +
  geom_line(data = df.fit, aes(x = age, y = pi)) +
  geom_line(data = df, aes(x = x, y = pi), col = "blue")

df.fit.d <- fit %>% 
  spread_draws(pred_h[i]) %>% 
  median_qi() %>% 
  mutate(age = i-1)

```

