---
title: "Survival kit for survival analyses"
description: |
  Deriving main mathematical expressions, simulating survival data applying an inverse CDF sampling with STAN and estimating a Piecewise Constant Hazards model.
author:
  - name: Benjamin Schlüter
date: 02-10-2022
output:
  distill::distill_article:
    self_contained: false
    highlight: monochrome
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

In this post, I would like to introduce three complementary mathematical functions frequently encountered in survival analyses: the hazard, the survival and the age at death probability density function. Using STAN, we will then simulate survival data using the inverse cumulative density function (CDF) sampling method assuming the hazard has a Gompertz form. We conclude by estimating the hazard rate on our simulated dataset with a Piecewise Constant Hazard (PCH) model fitted with STAN.

# Complementary mathematical functions

Let's denote X: the age at death of an individual and assume that this random variable is $X\geq 0$ and continuous. 

Surviving beyond age x is expressed as

$$
\begin{align}
S(x) & = Pr(X>x) \\
S(x) & = 1 - F(x) \\
S'(x) & = -f(x) \\
\Rightarrow S(x) & = -\int^{\infty}_xf(t)dt
\end{align}
$$

Let's now express the hazard rate, also called the force of mortality:

$$
\begin{align}
h(x) &= lim_{\Delta x \rightarrow 0} \frac{Pr(x \leq X < x + \Delta x | X > x)}{\Delta x} \\
 &= lim_{\Delta x \rightarrow 0} \frac{Pr(X < x + \Delta x) - Pr(X \leq x)}{\Delta x} \cdot \frac{1}{Pr(X>x)} \\
 &= \frac{f(x)}{S(x)}.
\end{align}
$$

Thus, the force of mortality can be expressed as

$$
\begin{align}
\boxed{h(x) = -\frac{d}{dx}ln(S(x))}
\end{align}
$$

and re-arranging the above equation, we express the survival probability at age x as

$$\boxed{S(x) = e^{-\int^x_0h(t)dt} = e^{-H(x)}}$$

where $H(x)$ is called the cumulative hazard at age x.

Note that when $h(x)$ is constant, 

$$S(x) = e^{-hx}.$$

From earlier expression of $h(x)$, we can derive the age at death probability density function

$$\boxed{f(x) = h(x)S(x)}.$$

These formulas show that given either the hazard rate or the survival function, it is possible to get everything else.


# The Gompertz case

Assuming the hazard rate has the Gompertzian form

$$h(x) = ae^{bx}$$
we can easily derive the two other functions. We start by deriving the cumulative hazard function

$$
\begin{align}
H(x) & = \int^x_0h(t)dt \\
& = \int^x_0ae^{bt}dt \\
& = \frac{a}{b}[e^{bx} - 1]
\end{align}
$$

and replacing it in the survival function

$$
\begin{align}
S(x) & = e^{-H(x)} \\
& = e^{\frac{a}{b}[1 - e^{bx}]}.
\end{align}
$$

It is then straightforward to obtain the death probability density function by multiplying both


$$
\begin{align}
f(x) & = h(x)S(x) \\
& = ae^{bx} e^{\frac{a}{b}[1 - e^{bx}]} 
\end{align}
$$


Suppose now that we follow individuals from their 55$^{th}$ birthday to their death. Using the [Human Mortality Database for Belgium](https://www.mortality.org/cgi-bin/hmd/country.php?cntr=BEL&level=1) in 2015, we estimated the Gompertz parameters to be $a = 0.0006$ and $b = 0.128$ on the ages 55 to 95 years old. Note that we restricted the estimation to these ages as the Gompertz model does not perform well for the youngest and oldest ages.

With this in hand, we can now plot our three complementary functions assuming Gompertz hazard and setting $a$ and $b$ to the values we estimated.



```{r a_b, include = FALSE}

library(ggplot2)
library(tidyverse)
library(rstan)
library(tidybayes)
library(cowplot)

source("./modeling/function/theme_dark_blue.R")

# Gompertz parameters
a <- 0.0006
b <- 0.128
# Ages
ages <- 55:120

# Functions
# Hazard rate
h <- function(a, b, x){a*exp(b*x)}
# Survival function
S <- function(a, b, x){exp((a/b)*(1 - exp(b*x)))}

# Create data set
df <- tibble(age = ages,
             x = ages - 55, # start at age 55
             haz = h(a, b, x),
             Surv = S(a, b, x),
             pdf = haz * Surv)

```


```{r plot_truth, echo= FALSE, preview = TRUE}
fig.h <- ggplot(df, aes(x = ages, y = haz)) +
  geom_line(col = "skyblue3",
            size = 1.2) +
  theme_dark_blue() +
  labs(y = "h(x)",
       x = "Age")

fig.logh <- ggplot(df, aes(x = ages, y = log(haz))) +
  geom_line(col = "skyblue3",
            size = 1.2) +
  theme_dark_blue() +
  labs(y = "ln(h(x))",
       x = "Age")

fig.S <- ggplot(df, aes(x = ages, y = Surv)) +
  geom_line(col = "skyblue3",
            size = 1.2) +
  theme_dark_blue() +
  labs(y = "S(x)",
       x = "Age")

fig.pi <- ggplot(df, aes(x = ages, y = pdf)) +
  geom_line(col = "skyblue3",
            size = 1.2) +
  theme_dark_blue() +
  labs(y = "f(x)",
       x = "Age")

plot_grid(fig.h,
          fig.logh,
          fig.S,
          fig.pi,
          ncol = 2,
          nrow = 2)
```

# Simulating survival data

The inverse CDF sampling method can be used to generate data. This method uses the fact that

$$ U = F(X) \sim U_{[0,1]}$$

Therefore, if $u_1,...,u_n$ is an independent and identically distributed (iid) random sample from $U_{[0,1]}$, then

$$\{x_i:x_i = F^{-1}(u_i), i=1,...,n\}$$
is an iid random sample from $F()$. The algorithm depends on the feasibility to inverse $F()$. 

Applying this method to the survival function, we can simulate ages at death


$$
\begin{align}
F(x_i) & = u_i \\
S(x_i) & = 1- u_i \\
e^{\frac{a}{b}[1 - e^{bx_i}]} &  = 1-u_i \\
1 - e^{bx_i} & = \frac{b}{a}log(1-u_i)  \\
x_i & = \frac{1}{b}log(1 - \frac{b}{a}log(1-u_i))
\end{align}
$$

with $u_i \sim U_{[0,1]}$.

The corresponding STAN code is


```{r code_STAN}
writeLines(readLines("./modeling/stan/simu_gomp.stan"))

```

The simulation could also easily be performed with base R with the following code

```{r simu_R, echo = TRUE, eval = FALSE}
u <- runif(N, 0, 1)
t <- (1/b)*log(1-((b/a)*log(1-u)))
```



```{r simu_stan, echo = FALSE, message = FALSE, results = "hide"}
# sample size
N <- 2e4

stan_data <- list("N" = N,
                  "a" = a,
                  "b" = b)

simu <- stan(file="./modeling/stan/simu_gomp.stan", data=stan_data,
             iter=1, chains=1, seed=411991,
             algorithm="Fixed_param")

t <- array(extract(simu)$t[1,])
# range(t)
```


We can now compare the density of the simulated ages at death with the true density to check if our simulation worked fine.  

```{r fig_density_simu, echo = FALSE}
# compare histogram with true pdf of t
ggplot() +
  geom_density(data = tibble(x = t + 55), 
                 aes(x = x, col = "f(x) from simulation"),
               size = 1.2) +
  geom_line(data = df, 
            aes(x = age, y = pdf, col = "true f(x)"),
            size = 1.5) +
  theme_dark_blue() +
  theme(legend.position = c(0.2, 0.85)) +
  scale_color_manual(values = c("true f(x)" = "skyblue3",
                                "f(x) from simulation" = "white")) +
  labs(y = "Density",
       x = "Age")
```



# Hazard estimation with a PCH model


Suppose now that we observe the simulated ages at death, not knowing the true hazard rate. Despite the fact that age at death is continuous, we can divide the time into 1-year intervals and assume that the hazard is constant in each interval. We can then use a Piecewise Constant Hazards model to estimate the hazard in these intervals. 

The following code transforms the simulated age at death in continuous time, `t`, into counts of deaths and exposures at each age interval.

```{r counts, echo = TRUE}
# width of interval (1 year)
eps <- 1
# containers
dth <- c()
exp <- c()
# breaks for 1-year age interval
brks <- seq(0, ceiling(max(t)), eps)
for (i in 1:(length(brks)-1)){
    # death counts by interval
    d <- sum( (t >= brks[i] & t < brks[i+1]) )
    
    # time lived in interval by those dying
    e_d <- t[t >= brks[i] & t < brks[i+1]] - brks[i]
    # time lived in interval by those surviving
    e <- sum(t >= brks[i+1])*eps + sum(e_d)
    
    dth <- c(dth, d)
    exp <- c(exp, e)
}
# create data set
df.counts <- tibble(interval = head(brks, -1),
               mid = head(brks, -1) + (eps/2),
               age = mid + 55,
               dth = dth,
               exp = exp)
```


These counts data are plotted on the bar charts below

```{r fig_count, echo = FALSE}

fig.dth <- ggplot(df.counts) +
geom_col(aes(x = age, y = dth),
         col = "black",
                 fill = "white") +
  theme_dark_blue() +
  labs(y = "Death counts",
       x = "Age")

fig.exp <- ggplot(df.counts) +
geom_col(aes(x = age, y = exp),
         col = "black",
                 fill = "white") +
  theme_dark_blue() +
  labs(y = "Exposure",
       x = "Age")

plot_grid(fig.dth,
          fig.exp,
          ncol = 2)
```

Assuming that the hazard is constant in each interval

$$
h(x) = \lambda_x ~ \forall x \in [x, x+1) 
$$

we can write the PCH model as follows

$$
D_x \sim Poisson(\lambda_xE_x)
$$
where $D_x$, $E_x$ and $\lambda_x$ are death counts, exposure and hazard rate for age $x$, respectively. The following STAN code estimates the model


```{r stan_data, echo = FALSE, results = "hide", message = FALSE}
 # Poisson estimation with STAN
stan_data <- list(
  "A" = dim(df.counts)[1],
  "age" = df.counts$mid,
  "n" = df.counts$exp,
  "d" = df.counts$dth
  )

options(mc.cores = parallel::detectCores()-1)


fit = stan("./modeling/stan/poi_for_gomp.stan",
           iter = 8000,
           chains = 4,
           data = stan_data)

```


```{r pch_stan}
writeLines(readLines("./modeling/stan/poi_for_gomp.stan"))
```

Finally, we extract the posterior draws of $\lambda_x$ and use them to get posterior draws of $S(x)$ and $f(x)$ using the math expressions presented earlier. Looking at the figures below, the PCH model is close to the truth with $N=2e4$ simulated ages at death.

```{r post_draws, echo = FALSE}
# Obtain S and pi from estimated mx
h.hat <- exp(extract(fit)$eta)
S.hat = sapply( 1:dim(h.hat)[1], function(x) exp(-cumsum(h.hat[x, ])) )
S.hat <- t(S.hat)
pi.hat = h.hat*S.hat

df.hat <- tibble(x = head(brks, -1),
                 age = x + 55, 
             h = apply(h.hat, 2, median),
             h.low = apply(h.hat, 2, quantile, probs = 0.005),
             h.up = apply(h.hat, 2, quantile, probs = 0.995),
             S = apply(S.hat, 2, median),
             S.low = apply(S.hat, 2, quantile, probs = 0.005),
             S.up = apply(S.hat, 2, quantile, probs = 0.995),
             pi = apply(pi.hat, 2, median),
             pi.low = apply(pi.hat, 2, quantile, probs = 0.005),
             pi.up = apply(pi.hat, 2, quantile, probs = 0.995))

# Compare h with true
fig.h.fit <- ggplot() +
  geom_line(data = df.hat, aes(x = age, y = h,
            col = "posterior PCH")) +
  geom_ribbon(data = df.hat, aes(x = age, ymin = h.low, ymax = h.up, col = "posterior PCH", fill = "posterior PCH"),
              alpha = 0.3) +
  geom_line(data = df %>% filter(age<= max(df.hat$age)), aes(x = age, y = haz, col = "truth"), 
            size = 1,
            linetype = "dashed") +
  scale_color_manual(values = c("posterior PCH" = "red4",
                                "truth" = "skyblue3")) +
  scale_fill_manual(values = c("posterior PCH" = "red4",
                                "truth" = "skyblue3")) +
  theme_dark_blue() +
  theme(legend.position = "none") +
  labs(y = "h(x)",
       x = "Age") 

fig.log.h.fit <- ggplot() +
  geom_line(data = df.hat, aes(x = age, y = log(h),
            col = "posterior PCH")) +
  geom_ribbon(data = df.hat, aes(x = age, ymin = log(h.low), ymax = log(h.up), col = "posterior PCH", fill = "posterior PCH"),
              alpha = 0.3) +
  geom_line(data = df %>% filter(age<= max(df.hat$age)), aes(x = age, y = log(haz), col = "truth"), 
            size = 1,
            linetype = "dashed") +
  scale_color_manual(values = c("posterior PCH" = "red4",
                                "truth" = "skyblue3")) +
  scale_fill_manual(values = c("posterior PCH" = "red4",
                                "truth" = "skyblue3")) +
  theme_dark_blue() +
  theme(legend.position = "none") +
  labs(y = "ln(h(x))",
       x = "Age") 


# Compare S with true
fig.S.fit <- ggplot() +
  geom_line(data = df.hat, aes(x = age, y = S,
            col = "posterior PCH")) +
  geom_ribbon(data = df.hat, aes(x = age, ymin = S.low, ymax = S.up, col = "posterior PCH", fill = "posterior PCH",),
              alpha = 0.4) +
  geom_line(data = df, aes(x = age, y = Surv, col = "truth"), 
            size = 1,
            linetype = "dashed") +
  scale_color_manual(values = c("posterior PCH" = "red4",
                                "truth" = "skyblue3")) +
  scale_fill_manual(values = c("posterior PCH" = "red4",
                                "truth" = "skyblue3")) +
  theme_dark_blue() +
  theme(legend.position = "none") +
  labs(y = "S(x)",
       x = "Age") +
  annotate("segment", x = c(92, 92), xend = c(96, 96), 
                 y = c(0.7, 0.9), yend = c(0.7, 0.9), 
                 size = 2, color = c("red4", "skyblue3")) +
  annotate("text", label = c("fit & 99% CI", "truth"),
           x = c(112, 112), y = c(0.7, 0.9), size = 3.5, color = c("red4", "skyblue3"))

# Compare pdf with true
fig.pi.fit <- ggplot() +
  geom_line(data = df.hat, aes(x = age, y = pi,
            col = "posterior PCH")) +
  geom_ribbon(data = df.hat, aes(x = age, ymin = pi.low, ymax = pi.up, col = "posterior PCH", fill = "posterior PCH",),
              alpha = 0.4) +
  geom_line(data = df, aes(x = age, y = pdf, col = "truth"), 
            size = 1,
            linetype = "dashed") +
  scale_color_manual(values = c("posterior PCH" = "red4",
                                "truth" = "skyblue3")) +
  scale_fill_manual(values = c("posterior PCH" = "red4",
                                "truth" = "skyblue3")) +
  theme_dark_blue() +
  theme(legend.position = "none") +
  labs(y = "f(x)",
       x = "Age")

plot_grid(fig.h.fit,
          fig.log.h.fit,
          fig.S.fit,
          fig.pi.fit,
          ncol = 2) 

```


# Summary

Survival analyses mainly require three complementary mathematical expressions: the hazard rate, the survival function and the age at death probability density function. Once we have a given mathematical form for the hazard rate (ie Gompertz), the two other functions can be derived. In order to test models, the inverse CDF sampling method allows to simulate survival data. One model considered in this post is the PCH model that can be modelled in a Bayesian estimation framework with STAN. Note that we did not consider censoring or truncation but it will be addressed in a future post.
