---
title: "Maximum likelihood estimation of mortality models with Newton-Raphson algorithm"
description: |
  Estimation of the Gompertz and Log-quadratic models of mortality with maximum likelihood by applying Newton-Raphson maximization algorithm
author:
  - name: Benjamin Schlüter
date: 10-19-2021
output:
  distill::distill_article:
    self_contained: false
---

This blog post shows how to estimate parameters of a mortality model by maximum likelihood instead of using regression. The aim is to show the influence of outliers at the highest age on the parameters.Regression cannot estimates mx=0 while this is not a problem for likelihoo estimation. Optim() does not work and no real control of what is happening.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo = FALSE}
library(tidyverse)
library(viridis)
library(scales)
```


We'll study old age mortality, higher or equal than 80 years old.

```{r tidy data, echo = FALSE}
dth = read.delim("./modeling/data/dthES_1x1.txt", 
                 skip = 1,
                 sep = "", 
                 stringsAsFactors=FALSE,
                 header = TRUE)

exp = read.delim("./modeling/data/expES_1x1.txt", 
                 skip = 1,
                 sep = "", 
                 stringsAsFactors=FALSE,
                 header = TRUE)

df.dth = dth %>% 
  filter(Year == 2018) %>% select(Year, Age, Female) %>% 
  mutate(Female = as.numeric(Female))

df.exp = exp %>% 
  filter(Year == 2018) %>% select(Year, Age, Female)

df = df.dth %>% 
  left_join(df.exp, 
            by = c("Year", "Age")) %>% 
  rename("D" = Female.x,
         "N" = Female.y) %>% 
  mutate(mx = D/N,
         Age = ifelse(Age == "110+", 110, Age),
         Age = as.numeric(Age),
         Age2 = Age^2)

ggplot(df, aes(x = Age, y = log(mx))) +
  geom_point(data = df %>% filter(Age >= 80), 
             size = 2, col = "#802582FF") +
  geom_point(data = df %>% filter(Age < 80), 
             size = 2, col = "#802582FF",
             alpha = 0.2) +
  geom_rect(xmin = 79, xmax=111, ymin = -5, ymax = 0.1,
            alpha = 0.1, col = "#238A8DFF", fill = NA,
            size = 2) + 
  theme_bw() +
  theme(plot.title = element_text(size = 16, face = "bold",hjust = 0.5),
        panel.grid.major.y = element_line(linetype = "dashed"),
        panel.grid.minor.y = element_line(linetype = "dashed"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.border = element_blank(),
        axis.line.x = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.line.y = element_line(size = 0.5, linetype = "solid", colour = "black")) +
  annotate("text",
           label = "Old age \n mortality",
            x = 95,
            y = -3.5,
            col = "#238A8DFF") +
  labs(title = "Mortality's Age Schedule",
       y = expression(paste("log(", {}[1], m[x], ")" )))
```


```{r regression estimation, echo = FALSE}
# remove young age mortality
df = df %>% filter(Age >= 80)

fit.reg = lm(log(mx) ~ 1 + Age, 
             df)
coef.reg = fit.reg$coefficients

```

Newton-Raphson's maximization algorithm.

```{r NR gompertz estimation, echo = TRUE}
# Design matrix
X = matrix(c(rep(1, dim(df)[1]), df$Age),
           ncol = 2)
# vector of deaths and exposure
D = df$D
N = df$N
# Expected deaths according to
# alpha
Dhat = function(alpha) {
  lambda.hat = X %*% alpha
  return( as.numeric( N * exp(lambda.hat)))
}
# NR algorithm
next_alpha = function(alpha) {
  dhat = Dhat(alpha)
  M = solve ( t(X) %*% diag(dhat) %*% X)
  v = t(X) %*% (D - dhat)
  return( alpha + M %*% v)
}

a = matrix(0, 2, 15)
for (i in 2:ncol(a)) { a[,i] = next_alpha(a[,i-1])}

```

Alpha values over N-R iterations. Convergence before tenth iteration.

```{r plot iterations NR gompertz, echo = FALSE}
df.a = t(a) %>% 
  as.data.frame() %>% 
  rename("alpha1" = V1,
         "alpha2" = V2)

ggplot(df.a, aes(x = 1:15)) +
  geom_line(aes(y = alpha1, col = "alpha1"), size = 1.2) +
  geom_line(aes(y = alpha2, col = "alpha2"), size = 1.2) +
  theme_bw() +
  theme(plot.title = element_text(size = 16, face = "bold",hjust = 0.5),
        legend.position = c(0.7, 0.55),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.direction = "horizontal",
        panel.grid.major.y = element_line(linetype = "dashed"),
        panel.grid.minor.y = element_line(linetype = "dashed"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.border = element_blank(),
        axis.line.x = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.line.y = element_line(size = 0.5, linetype = "solid", colour = "black")) +
  scale_color_manual(values = c("alpha1" = "#C7B76BFF",
                                "alpha2" = "#00336FFF"),
                     labels = c(expression(alpha[1]), expression(alpha[2]))) +
  labs(title = "Alpha parameters over N-R iterations",
       y = "Values",
        x = "Iterations")

```


Gompertz fit of old age mortality. Regression gives more weights to rates at oldest ages.

```{r plot outputs NR gompertz, echo = FALSE}
fitted.logmx = X %*% a[, ncol(a)]

df = df %>% 
  mutate(log.mx.hat.NR.gptz = fitted.logmx,
         log.mx.hat.reg.gptz = coef.reg[1] + coef.reg[2]*Age)

ggplot(df, aes(x = Age, y = log(mx), group = 1)) +
  geom_point(size = 2, col = "#802582FF") +  
  geom_line(aes(x = Age, y = log.mx.hat.NR.gptz, 
                col = "Gompertz by N-R" ),
            size = 1.2) +
  geom_line(aes(x = Age, y = log.mx.hat.reg.gptz, 
                col = "Gompertz by regression" ),
            size = 1.2) +
  theme_bw() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        legend.position = c(0.8, 0.2),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        panel.grid.major.y = element_line(linetype = "dashed"),
        panel.grid.minor.y = element_line(linetype = "dashed"),
        panel.grid.major.x = element_line(linetype = "dashed"),
        panel.grid.minor.x = element_line(linetype = "dashed"),
        panel.border = element_blank(),
        axis.line.x = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.line.y = element_line(size = 0.5, linetype = "solid", colour = "black")) +
  scale_color_manual(values = c("Gompertz by N-R" = "#D6456CFF",
                                "Gompertz by regression" = "#3F4788FF")) +
  labs(title = "Gompertz fits",
       y = expression(paste("log(", {}[1], m[x], ")" )))
```


```{r likelihood estimation gompertz, echo = FALSE, eval = FALSE}
pois.lik.gmptz <- function(pars, data){
  alpha = pars[1]
  beta = pars[2]
  ll = with(data, sum( D*(alpha+beta*Age) - N*exp(alpha+beta*Age) ) )
  return(-ll)
}

fit.pars = optim(c(coef.reg[1], coef.reg[2]), pois.lik.gmptz, 
      data=df, 
      method="BFGS")
# Comparison of reg, optim() and NR
ggplot(df, 
       aes(x = Age, y = log(mx), group = 1)) +
  geom_line() +
  geom_line(aes(x = Age, y = log.mx.hat.NR.gptz ),
              col = "blue") +
  geom_abline(slope = fit.pars$par[2],
              intercept = fit.pars$par[1],
              col = "navy") +
  geom_abline(slope = coef.reg[2],
              intercept = coef.reg[1],
              col = "red") +
  theme_bw()

```

Algorithm only need a third column in design matrix and an additional line in a. Alpha values over N-R iterations. Convergence before tenth iteration.


```{r NR log-quad estimation, echo = FALSE}
# Augment design matrix
X = matrix(c(rep(1, dim(df)[1]), df$Age, df$Age^2),
           ncol = 3)

a = matrix(0, 3, 15)
for (i in 2:ncol(a)) { a[,i] = next_alpha(a[,i-1])}
# round(a, 4)

df.a = t(a) %>% 
  as.data.frame() %>% 
  rename("alpha1" = V1,
         "alpha2" = V2,
         "alpha3" = V3)

ggplot(df.a, aes(x = 1:15)) +
  geom_line(aes(y = alpha1, col = "alpha1"), size = 1.2) +
  geom_line(aes(y = alpha2, col = "alpha2"), size = 1.2) +
  geom_line(aes(y = alpha3, col = "alpha3"), size = 1.2) +

  theme_bw() +
  theme(plot.title = element_text(size = 16, face = "bold",hjust = 0.5),
        legend.position = c(0.7, 0.55),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.direction = "horizontal",
        panel.grid.major.y = element_line(linetype = "dashed"),
        panel.grid.minor.y = element_line(linetype = "dashed"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.border = element_blank(),
        axis.line.x = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.line.y = element_line(size = 0.5, linetype = "solid", colour = "black")) +
  scale_color_manual(values = c("alpha1" = "#C7B76BFF",
                                "alpha2" = "#00336FFF",
                                "alpha3" = "#56C667FF"),
                     labels = c(expression(alpha[1]), 
                                expression(alpha[2]),
                                expression(alpha[3]))) +
  labs(title = "Alpha parameters over N-R iterations",
       y = "Values",
        x = "Iterations")


```


Gompertz and Log-Quadratic fits. Regression outputs more sensitive to outlying value.

```{r plot outputs NR log-quad, echo = FALSE}

fitted.logmx = X %*% a[, ncol(a)]

fit.reg = lm(log(mx) ~ 1 + Age + Age2, 
             df)
coef.reg.lq = fit.reg$coefficients

df = df %>% 
  mutate(log.mx.hat.NR.lq = fitted.logmx,
         log.mx.hat.reg.lq = coef.reg.lq[1] + coef.reg.lq[2]*Age + coef.reg.lq[3]*Age2)

ggplot(df, aes(x = Age, y = log(mx), group = 1)) +
  geom_point(size = 2, col = "#802582FF") +  
  geom_line(aes(x = Age, y = log.mx.hat.NR.gptz, 
                col = "Gompertz by N-R" ),
            size = 1.2) +
  geom_line(aes(x = Age, y = log.mx.hat.NR.lq, 
                col = "Log-Quadratic by N-R" ),
            size = 1.2) +
  geom_line(aes(x = Age, y = log.mx.hat.reg.lq, 
                col = "Log-Quadratic by regression" ),
            size = 1.2) +
  theme_bw() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        legend.position = c(0.8, 0.2),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        panel.grid.major.y = element_line(linetype = "dashed"),
        panel.grid.minor.y = element_line(linetype = "dashed"),
        panel.grid.major.x = element_line(linetype = "dashed"),
        panel.grid.minor.x = element_line(linetype = "dashed"),
        panel.border = element_blank(),
        axis.line.x = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.line.y = element_line(size = 0.5, linetype = "solid", colour = "black")) +
  scale_color_manual(values = c("Gompertz by N-R" = "#D6456CFF",
                                "Log-Quadratic by N-R" = "#74D055FF",
                                "Log-Quadratic by regression" = "#6B6C71FF")) +
  labs(title = "Gompertz and Log-Quadratic fits",
       y = expression(paste("log(", {}[1], m[x], ")" )))

```



```{r likelihood estimation log-quad, echo = FALSE, eval = FALSE}

# DOES NOT WORK
pois.lik.lq <- function(pars, data){
  alpha = pars[1]
  beta1 = pars[2]
  beta2 = pars[3]
  ll = with(data, sum( D*(alpha+beta1*Age+beta2*Age2) - N*exp(alpha+beta1*Age+beta2*Age2) ) )
  return(-ll)
}
 
fit.reg = lm(log(mx) ~ 1 + Age + Age2, 
             df)
coef.reg.lq = fit.reg$coefficients

df = df %>% 
  mutate(mx_lq_rg = coef.reg.lq[1] + coef.reg.lq[2]*Age + coef.reg.lq[3]*Age2)

ggplot(df, aes(x = Age, y = log(mx), group = 1)) +
  geom_line() +
  geom_line(aes(x = Age, y = mx_lq_rg)) +
  # geom_abline(slope = fit.pars$par[2],
  #             intercept = fit.pars$par[1],
  #             col = "blue") +
  geom_abline(slope = coef.reg[2],
              intercept = coef.reg[1],
              col = "red") +
  theme_bw()


fit.pars = optim(coef.reg.lq, pois.lik.lq, 
      data=df, 
      method="BFGS")

```


